---
title: "mtj_plots"
author: "Trisha Shukla"
date: "2025-02-28"
output: html_document
---

```{r}
# import necessary libraries 
library(readr)
library(tidyverse)
library(dplyr)
library(ggplot2)
```

```{r}
mtj <- read_csv("mtj_py.csv") # loading the dataset into R
view(mtj) # viewing the table 
```

```{r}
# filtering the data into three different objects, each of which is categorised based on one aromatic residue in aro_res (phenylalanine, tyrosine, tryptophan)

phe <- mtj %>% 
  filter(str_detect(aro_res, "F"))
phe

tyr <- mtj %>% 
  filter(str_detect(aro_res, "Y"))
tyr

trp <- mtj %>% 
  filter(str_detect(aro_res, "W"))
trp

```

```{r}
# counting the total number of residues for each aromatic residue group (F, Y, W) 

count_phe <- nrow(phe)
count_tyr <- nrow(tyr)
count_trp <- nrow(trp)

# combining into single df 

freq_aro <- data.frame(
  group = c("PHE", "TYR", "TRP"), 
  count = c(count_phe, count_tyr, count_trp)
)

# reordering the data 
freq_aro$group <- factor(freq_aro$group, levels = c("PHE", "TYR", "TRP"))

# plotting a bar plot to represent frequency of each aromatic residue 

p1 <- ggplot(freq_aro, aes(x = group, y = count, fill = group)) + 
  geom_bar(stat = "identity") + 
  labs(
    x = "", y = "Number of MTJs"
  ) + 
  theme_classic() + 
  scale_fill_manual(values = c("PHE" = "royalblue2", "TYR" = "firebrick2", "TRP" = "seagreen2")) + 
  theme(
    legend.title = element_blank(), 
    legend.position = "right", 
    axis.text.x = element_blank(), 
    axis.ticks.x = element_blank(), 
    axis.line = element_line(size = 0.5, color = "black")
  ) + 
  scale_y_continuous(
    breaks = seq(0, max(freq_aro$count), by = 45)
  )
  
p1 
# saving the plot 
ggsave("freq_mtj.png", plot = p1)
```

Next, a plot showing the distribution of chi1 and chi2 angles was created: 

```{r}
mtj_fixed <- mtj %>% 
  mutate(res_type = substr(aro_res, 1,1)) # creates an additional column with residue type listed specifically as F, W, or Y. 

head(mtj_fixed) 

install.packages("openxlsx")
library(openxlsx)
openxlsx::write.xlsx(mtj_fixed, "mtj_fixed.xlsx", rowNames=TRUE)
```

```{r}
# now we plot !!! 

mtj_fixed$res_type <- factor(mtj_fixed$res_type, levels = c("F", "W", "Y"))

library(ggh4x)

p <- ggplot(mtj_fixed, aes(x = chi1, y = chi2, colour = res_type)) +
  geom_point(size = 0.8) +
  labs(x = "chi2(ยบ)", y = "chi1(ยบ)") +
  theme(axis.line = element_line(colour = "black")) +
  coord_axes_inside(labels_inside = TRUE) + 
  scale_x_continuous(breaks = c(-80,-60,-40,-20, 20, 40, 60, 80)) + 
  scale_y_continuous(breaks = c(-120, -90,-60,-30,30,60,90,120)) +
  scale_color_manual(values = c("F" = "royalblue2","W" = "seagreen2", "Y" = "firebrick2")) + 
  guides(color = guide_legend(title = NULL, override.aes = list(shape = 16, size = 4))) + 
  theme_classic()
p
# saving the plot 
ggsave("angles_distribution.png", plot = p)
```

The next plot would be the distribution of distances (creating a density plot or histogram for the three different residues?)

```{r}
library(ggplot2)

p2 <- ggplot(mtj_fixed, aes(x = distance, fill = res_type)) + 
  geom_density(alpha = 0.5) + 
  labs(x = "Distance (ร)", y = "Density") + 
  scale_x_continuous(breaks = seq(1, 6, by = 0.25)) +
  scale_fill_manual(values = c("F" = "royalblue2", "Y" = "firebrick2", "W" = "seagreen2")) + 
   guides(fill = guide_legend(title = NULL)) + 
   theme_classic() 
p2

# saving the plot 
ggsave("distance_mtj.png", plot = p2)
```
Furthermore, the dataset was modified to contain the organism name for each accession ID: 

```{r}
mtj_edited <- read_csv("mtj_fixed_1.csv") # importing the modified dataset containing passenger domain type for each ID
mtj_edited
```
```{r}
bacteria <- read_csv("organism_list.csv") # importing names of organisms associated with each ID
bacteria
```

```{r}
mtj_final <- merge(mtj_edited, bacteria, by = "accession_ID", all.x = TRUE) # merges mtj_edited with bacteria based on accession ID 

view(mtj_final)
```

```{r}
nrow(mtj)
nrow(mtj_final)
```
```{r}
library(dplyr)

mean_mtj_per_PD <- mtj_final %>% 
  group_by(PD_type, accession_ID) %>%
  summarise(pair_count = n()) %>%
  group_by(PD_type) %>%
  summarise(mean_mtj = mean(pair_count))

print(mean_mtj_per_PD)
```
```{r}
library(ggpattern) # loading required package

mean_mtj_per_PD$fill_color <- ifelse(mean_mtj_per_PD$PD_type == "TcfA-TynE", "goldenrod2", "lightgray")

pd <- ggplot(mean_mtj_per_PD, aes(x = mean_mtj, y = PD_type, fill = fill_color)) + 
  geom_bar(stat = "identity", width = 0.7, size = 1, show.legend = FALSE, color = "black") +
  scale_x_continuous(expand = c(0, 0), breaks = seq(0, 10, by = 0.5), limits = c(0, 10), labels = function(x) ifelse(x %% 1 == 0, as.character(x), paste0(x))) +
  labs(x = "Frequency of MTJs per AT", y = "Passenger Domain Type") + 
  scale_fill_identity() +
  theme_classic() 

pd

ggsave("pd_freq.png", plot = pd, width = 8, height = 6, dpi = 300)
```

Now we do this whole thing all over again, but for beta-barrel domain function/type: 

```{r}
at_function <- read_csv("beta_barrel_function.csv") # importing names of organisms associated with each ID
at_function

mtj_at <- merge(mtj_final, at_function, by = "accession_ID", all.x = TRUE) 

mtj_at <- mtj_at %>% rename (p_annotation = `description/protein_annotation`)
view(mtj_at)
```

```{r}
library(dplyr)

mean_mtj_per_pa <- mtj_at %>% 
  group_by(p_annotation, accession_ID) %>%
  summarise(pair_count = n()) %>%
  group_by(p_annotation) %>%
  summarise(av_mtj = mean(pair_count))

print(mean_mtj_per_pa)
```

```{r}
library(ggpattern) # loading required package

rm(mtj_beta)
rm(mean_mtj_per_bb)
range(mean_mtj_per_pa$av_mtj)
```

```{r}

mean_mtj_per_pa$fill_color <- ifelse(mean_mtj_per_pa$p_annotation %in% c("autotransporter YapB", "autotransporter adhesin EhaB"), "chosen", "other")

p4 <- ggplot(mean_mtj_per_pa, aes(x = av_mtj, y = p_annotation, fill = fill_color)) + 
  geom_bar(stat = "identity", width = 0.7, size = 0.5, show.legend = FALSE, color = "black") +
  scale_x_continuous(expand = c(0, 0), breaks = seq(0, 10, by = 0.5),limits = c(0, 10), labels = function(x) ifelse(x %% 1 == 0, as.character(x), paste0(x))) +
  scale_fill_manual(values = c("chosen" = "goldenrod2", "other" = "lightgray")) +
  labs(x = "Frequency of MTJs per AT", y = "Protein annotation") + 
  theme_classic() + 
  theme(
    axis.text.y = element_text(size = 5, angle = 0, hjust = 1, vjust = 0.5, margin = margin(r = 5)) # Add margin to y-axis text
  )

p4

ggsave("at_type_freq.png", plot = p4, width = 8, height = 6, dpi = 300)
```

The next plot was created to visualise the frequency of aromatic residues on each beta-barrel strand from 1 to 12: 

```{r}
# plotting distribution of aromatic residue across the 12 strands, differentiated based on type of residue 

library(ggpattern)

strands <- ggplot(mtj_at, aes(x = factor(n_strand_aro), fill = res_type)) + 
  geom_bar_pattern(stat = "count", position = "dodge", pattern_density = 0.1, pattern_fill = "white", width = 0.7, color = "black", size = 0.5) + 
  labs(x = "Beta Strand Number", y = "Number of Aromatic Residues") + 
  scale_fill_manual(values = c("F" = "royalblue2", "W" = "seagreen2","Y" = "firebrick2")) + 
  theme_classic() + 
  theme(
    axis.text.x = 
      element_text(size = 10, angle = 0, hjust = 0.5), 
    axis.text.y = element_text(size = 10)
  ) + 
  guides(fill = guide_legend(title = NULL))
strands

ggsave("strands.png", plot = strands, width = 8, height = 6, dpi = 300)
```

Visualising plot(s) based on genus/species: 

```{r}
library(dplyr)
library(tidyr)

# extracting the genus from the organism column 
mtj_organism <- mtj_at %>% 
  mutate(genus = sapply(strsplit(as.character(organism), " "), `[`, 1))

# count the number of aromatic residues (aro_res) for each accession id per genus

count_mtj <- mtj_organism %>% 
  group_by(genus, accession_ID) %>% 
  summarise(aro_count = n())

# calculate the average number of MTJs per AT categorised by genus 

mean_count_by_genus <- count_mtj %>% 
  group_by(genus) %>% 
  summarise(mean_aro_count_per_AT = mean(aro_count, na.rm = TRUE))

mean_count_by_genus

# plot the results 

mean_count_by_genus$fill_color <- ifelse(mean_count_by_genus$genus == "Photorhabdus", "coral2", "lightgray") # differentiating the genus with highest number of MTJs by colour 

mean_mtj_genus <- ggplot(mean_count_by_genus, aes(y = genus, x = mean_aro_count_per_AT, fill = fill_color)) + 
  geom_bar(stat = "identity", show.legend = FALSE, color = "black", width = 0.7) + 
  labs(y = "Genus", x = "Mean Number of Aromatic Residues per AT") + 
  theme_classic() + 
  theme(axis.text.y = element_text(angle = 0, hjust = 1)) + 
  scale_fill_manual(values = c("coral2", "lightgray")) + 
  scale_x_continuous(expand = c(0, 0), breaks = seq(0, 10, by = 0.5),limits = c(0, 10), labels = function(x) ifelse(x %% 1 == 0, as.character(x), paste0(x)))

mean_mtj_genus
```











